\textbf{Introducing the Choice-Confidence (CHOCO) Model for Bimodal Scales Data: Application to the Relationship between Reality Beliefs about AI-Generated Faces and Beauty}

### 2025-06-11 paper/manuscript.tex
update

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, bottomrule=.15mm, colback=white, colframe=quarto-callout-note-color-frame, opacityback=0, breakable, arc=.35mm, leftrule=.75mm, toprule=.15mm, left=2mm]

@@

Despite significant advancements in psychological science following the
replication crisis (\citeproc{ref-open2015estimating}{Collaboration,
2015}), its progress is still hindered by its sub-optimal (or
inappropriate) usage of statistical tools
(\citeproc{ref-makowski2023we}{Makowski \& Waggoner, 2023}). A prevalent
issue is the continued reliance on linear models that assume normally
distributed (Gaussian) data - as this assumption often does not hold
true for many types of psychological outcomes. For instance, reaction
times typically exhibit skewed distributions, choices can be represented
as binary variables, and count data consists of strictly positive
integers. Applying models that presume normality and model the ``mean''
of the outcome variable can lead to misinterpretation and potentially
misleading conclusions when applied indiscriminately. It is thus
important that psychologists use models that can best describe (or
generate) the data they collect, to fully exploit them and bring more
nuance and accuracy to their conclusions.

@@

\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{./figures/fig1.png}

@@

Decision-making is often conceptualized as involving distinct processes:
the choice itself, and the confidence associated with that choice. In
experimental paradigms, these can be somewhat disentangled by prompting
participants to make a discrete choice selection (e.g., ``True''
vs.~``False''), followed by a separate confidence rating. However, this
artificial separation makes its joint analysis difficulty, and may not
reflect real-world decision-making, where individuals often express both
choice and confidence simultaneously using a single, continuous scale.
In such scales, each side can represent a distinct latent category, and
the distance from the midpoint can indicates the level of confidence or
certainty. This integrated response format typically results in bimodal
distributions, with peaks corresponding to the mean confidence on either
side. Traditional beta regression models, which assume unimodal
distributions within the (0,1) interval, are ill-suited for such data.
One alternative is to transform the data into two variables a
posteriori: binarizing the side to represent choice and calculating the
absolute distance from the midpoint to represent confidence. These can
then be modeled separately, for instance, using logistic regression for
choice and beta regression for confidence (see
\citeproc{ref-makowski2025too}{Makowski et al., 2025} for an example).
While this approach can provide additional insights into underlying
mechanisms compared to a unique model, it assumes psychological and
statistical independence between choice and confidence, which may not
hold true in practice.

@@

\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{./figures/fig2.png}

@@

\textbf{Study 1} aims to evaluate the CHOCO model's ability to better
capture subjective scale responses that (potentially) reflect an
underlying discrete choice, in comparison to existing models such as the
ZOIB and Beta-Gate. Specifically, we will assess whether 1) CHOCO
provides improved model fit, 2) yields deeper insights into
population-level effects than traditional approaches (gender differences
in reality beliefs), and 3) allows for the reliable estimation of
interpretable individual-level parameters through random effects.

To this end, we analyze data from two separate studies in which
participants judged whether a face image was AI-generated (``fake'') or
a real photograph. \textbf{Study 2} will apply this model to more subtle
effects, such as the effect perceived facial attractiveness and beauty
on reality judgments.

@@

\textbf{TODO}: Write some introduction about reality beliefs in the
context of ambiguity. Underline that what drives judgments and beliefs
of reality remains unclear. And this is important in today's post-truth
era.

In a post-truth era (\citeproc{ref-lewandowsky2017beyond}{Lewandowsky et
al., 2017}), where near-perfect simulations are increasingly feasible
due to rapid technological advancements, distinguishing between what is
real and what is fake poses a growing challenge {[}\textbf{REF}{]}. As
these technologies advance and physical cues become indistinguishable,
individuals are likely to rely more heavily on alternative sources of
information -such as contextual cues and cognitive heuristic -when
making reality judgments {[}i.e., simulation monitoring; Makowski
(\citeproc{ref-makowski2018cognitive}{2018}), Makowski, Sperduti, et al.
(\citeproc{ref-makowski2019phenomenal}{2019}){]}. This reliance is
particularly pronounced under conditions of high ambiguity, where
processing demands are higher and the atomization and
decontextualization of information (especially prevalent in online
environments) further complicate the discernment of authenticity.

Another unknown is whether these beliefs are fully stimuli-driven or
whether there are some stable individual tendencies that are at stake.
For instance, in the absence of clear contextual information,
individual-level characteristics may increasingly guide beliefs about
reality. Traits such as narcissism have been shown to influence such
judgments (\citeproc{ref-makowski2025too}{Makowski et al., 2025}).
Moreover, it has been proposed that the interaction between stimulus
characteritics and individual differences plays a crucial role in
shaping reality beliefs. For example, self-reported arousal has been
found to predict the likelihood of perceiving an image as real
(\citeproc{ref-azevedo2020body}{Azevedo et al., 2020}), while judging a
stimulus as fake has been associated with emotional downregulation
(\citeproc{ref-makowski2019phenomenal}{Makowski, Sperduti, et al.,
2019}). {[}\textbf{should probably add other examples}{]}

These findings underscore that what drives beliefs and judgments of
reality remains unclear. Yet, understanding these mechanisms is
increasingly vital given the proliferation of AI-generated content and
the growing risk of misjudging the authenticity of stimuli, particularly
in contexts with significant societal consequences (e.g., the influence
of misinformation on democratic processes).

@@

The sample is the same as in (\citeproc{ref-makowski2025too}{Makowski et
al., 2025}). We included all heterosexual and bisexual (these two groups
did not seem to differ based on preliminary analyses and were thus
grouped to maximize power) male and female participants, for a final
sample of 141 participants (Mean age = 28.4, SD = 9.0, range: {[}19,
66{]}; Sex: 47.5\% females). For each participant, we included only
stimuli of the opposite gender (i.e., all 89 female faces for men and 20
male faces for women).

@@

In the first phase, participants viewed 109 neutral-expression
photographs of faces (random order, display time of 500 ms) from the
American Multiracial Face Database
(\citeproc{ref-chen2021broadening}{Chen et al., 2021}). After each
image, participants rated the face on trustworthiness, familiarity,
attractiveness, and beauty using visual analog scales. In the second
phase, participants were informed that ``about half of the previously
seen images were AI-generated''. The same faces were presented again in
a new random order (same display time), followed by ratings of
``reality'' (whether they believed the image was fake - left anchor - or
a real - right anchor). Makowski et al.
(\citeproc{ref-makowski2025too}{2025}) showed that most stimuli yielded
highly variable reality ratings, suggesting that idiosyncratic cognitive
processes were at play.

@@

We fitted 3 models to predict the reality ratings: a ZOIB model, a
Beta-Gate model, and the CHOCO model. For all models and each parameter,
the full formula was entered:
\(Real\sim Sex + (1 | Participant) + (1 | Item)\) (with Sex as the main
predictor and participants and items entered as random intercepts). The
models were run using \emph{brms}
(\citeproc{ref-burkner2017brms}{Bürkner, 2017}) R package, and analyzed
using the \emph{easystats} collection of packages
(\citeproc{ref-easystats2022}{Lüdecke et al., 2022}). To maximize the
comparability across models We used the default priors (uniform) for all
models, and we ran 16 chains of 1400 iterations each on the University
of Sussex High-Performance Computing (HPC) cluster.

@@

Finally, we will run a correlation analysis of the models'
individual-level estimates against ``empirical'' (indices computed
directly on the observed data), including the empirical \emph{p}
(\(P(y > 0.5)\)), the overall \emph{conf} (\(mean(|y - 0.5|)\)),
\emph{pex} (\(P(y \in [0, 1])\)) and \emph{bex}
(\(P(y == 1) / P(y \in [0, 1])\)), assessing whether the model's
estimate are in-line with easily interpretable indices.

@@

{\caption{{TODO.}{\label{fig-three}}}}

@@

The models did converge without divergent transitions, and the effective
sample size was sufficient for all parameters (all
\(n_{\text{eff}} > 1000\)). The difference in predictive accuracy, as
indexed by Expected Log Predictive Density (ELPD-WAIC), suggests that
the \emph{CHOCO} is the best model (\(ELPD = -203.54\)), followed by
\emph{Beta-Gate} (\(\Delta_{ELPD} = -1794.57 \pm 63.12, p < .001)\) and
\emph{ZOIB} (\(\Delta_{ELPD} = -1833.59 \pm 63.52, p < .001\)). See
Figure~\ref{fig-three} for the posterior predictive checks, showing that
only the CHOCO model managed to capture the bimodal distribution of
data.

@@

The ZOIB model suggested that women had higher mean (\emph{mu}
parameter) scores of reality beliefs
(\(\Delta_{\text{Female}} = 0.20,~95\%~ CI~[0.03, 0.37], pd = 98.77\%\)),
less extreme values (\emph{zoi} parameter; STATS) but more ones relative
to zeros (\emph{coi} parameter; STATS).

The Beta-Gate model similarly suggested that women had higher mean
(\emph{mu} parameter) scores of reality beliefs (STATS), less extreme
values (\emph{pex} parameter; STATS), and a greater tendency to answer
one relative to zero (\emph{bex} parameter; STATS).

The CHOCO model shows that women had a higher probability \emph{p} of
judging faces as real (STATS), but are not more confident when doing so
(\emph{confright} parameter; STATS). However, they were less confident
when answering that an image was AI-generated (\emph{confleft}
parameter; STATS). There were also less likely to produce extreme
answers (\emph{pex} parameter; STATS), but no strong evidence supporting
a directional bias was observed (\emph{bex} parameter; STATS).

@@

{\caption{{TODO.}{\label{fig-four}}}}

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./figures/fig4.png}

@@

The CHOCO model yielded reliable estimates for all parameters except
\emph{bex} for participants (\emph{confright} D-vour = 0.94,
\emph{confleft} D-vour = 0.91, \emph{pex} D-vour = 0.79, \emph{p} D-vour
= 0.73, \emph{precright} D-vour = 0.77, \emph{precleft} D-vour = 0.67).
Item's variability was primarily reflected through the \emph{p}
parameter (D-vour = 0.86).

@@

Finally, we also show that the CHOCO model was able to capture reliable
and interpretable individual-level parameters, supporting its value to
measure inter-individual differences. An interesting dissociation
emerged between participant- and item-level variability: the latter
seemed mostly to be represented in the \emph{p} parameter, while
participants reliably varied in most of the components (aside from
\emph{bex}). This could be interpreted as that external item
characteristics primarily influence the probability of being judged as
real vs.~fake, while the expressed confidence is first and foremost an
individual characteristic.

@@

In this evolving landscape, understanding the cognitive mechanisms that
underpin our judgments of reality becomes paramount. Despite the
increasing prevalence of digitally altered or AI-generated content,
humans still rely on certain heuristics to assess authenticity. One such
heuristic might be facial attractiveness. Attractiveness appraisals are
known to be automatic and unconscious (\citeproc{ref-hou2023neural}{Hou
et al., 2023}; \citeproc{ref-hung2016unconscious}{Hung et al., 2016};
\citeproc{ref-luo2019robust}{Luo et al., 2019}), and carry strong
real-life consequences, as demonstrated by the large body of literature
on the ``beauty premium'' (\citeproc{ref-gulati2024beautiful}{Gulati et
al., 2024}; \citeproc{ref-kukkonen2024beauty}{Kukkonen et al., 2024};
\citeproc{ref-little2021facial}{Little, 2021};
\citeproc{ref-pandey2021face}{Pandey \& Zayas, 2021})\footnote{As an
  eloquent example, Monk Jr et al.
  (\citeproc{ref-monk2021beholding}{2021}) reported in a large
  representative US sample that the magnitude of earnings disparities
  among white women along the perceived attractiveness continuum exceeds
  in magnitude the canonical black-white race gap.}

\textbf{TODO:} However, it is important to allow for the possibility of
a U-shaped relationship. For instance, Peng et al.
(\citeproc{ref-peng2020faces}{2020}) reported that both attractive and
unattractive people sold more products (underlining a ``plainness
penalty'').

Although its role as a potential modulator of reality beliefs remains
underexplored, Makowski et al. (\citeproc{ref-makowski2025too}{2025})
found significant associations between participants' realness ratings
and facial salience. Female participants judged both highly attractive
and highly unattractive faces as more likely to be real, whereas for
male participants, this effect was limited to highly attractive faces.
These findings offer a complementary perspective to those of Miller et
al. (\citeproc{ref-miller2023ai}{2023}), who reported that AI-generated
faces tend to be rated as more attractive, and that participants used
attractiveness as a distinguishing cue between real and AI-generated
faces. Together, these results highlight a possible bidirectional and
context-dependent role of attractiveness in shaping reality judgments.

In this study, we will apply the CHOCO model to the data of MAKOWSKI,
and see whether different analysis approaches lead to different
conclusions. We will then try to replicate these findings in a new
sample of participants.

@@

The first sample includes the same participants as study 1 (see above).

\textbf{TODO} for sample 2.

The final sample included 189 participants (Mean age = 28.4, SD = 14.0,
range: {[}18, 69{]}; Sex: 76.2\% females).

@@

For the second sample, the procedure was relatively similar with a few
key differences. Most importantly, subjective ratings were collected
using a 7-point Likert scale ranging from 0 to 6, rather than a visual
analog scale, which included a clear midpoint. Additionally, the
familiarity rating was removed from the set of facial evaluations. In
the second phase, realness judgments were also recorded on a 7-point
Likert scale, this time ranging from -3 to +3, again with a defined
midpoint, replacing the visual analog format used in the first study.
For this scale, the endpoints were labeled `AI-generated' and
`Photograph', replacing the `Fake' and `Real' anchors used in Study 1.

There were also modifications to the instructions provided to
participants. In Phase 1, an additional prompt was introduced, stating
that the study was conducted in partnership with a young AI startup.
This was intended to strengthen participants' belief that the faces they
were shown could plausibly be either real or AI-generated. Additionally,
participants were explicitly asked to respond based on their immediate,
first impressions---an instruction not included in the first study.

In Phase 2, participants were not informed about the proportion of
AI-generated versus real faces encountered during Phase 1. This omission
was deliberate, aimed at reducing demand characteristics and preventing
participants from mechanically selecting half the faces as real and half
as AI-generated. Instead, they were encouraged to base their responses
solely on their level of confidence regarding the origin of each image.

@@

To compare the benefits of CHOCO models to a ``traditional'' analytic
approach, we started by fitting a frequentist linear mixed model to
predict reality beliefs with the formula
\(Real\sim Sex / poly(Attractive, 2) + (poly(Attractive, 2)|Participant) + (1|Item)\).
The 2nd degree raw polynomial term was included to allow for potentially
non-linear relationships (note that the first and second degree effects
of \emph{raw} polynomials can be interpreted independently as the linear
part and the ``curvy'' part of the relationship). For the CHOCO model,
mildly informative and effect-agnostic (i.e., centered at zero) priors
were used. The same formula was used for all parameters, except that
(based on the reliability analysis of Study 1), items were only included
as random effects for \emph{p}, and participants were not included for
\emph{bex}.

\textbf{TODO}. For sample 2, the analysis was based on that of Sample 1.
The main differences are 1) the inclusion of the ``Condition'' (whether
the picture was presented as ``Real'' of ``Fake'' in the first phase of
the experiment) as an additional predictor (it was entered as the only
random slope for all participant random effects); 2) the inclusion of an
additional parameters, \emph{pmid}, modelling the probability of
answering the middle-point of the Likert scale (representing an
``undecided'' option).

The same analysis was applied for Beauty.

@@

In sample 1, the traditional approach suggested a significant linear
relationship between the mean level of ``Reality'' and attractiveness
(\(\beta_{poly1}=3.42, ~95\%~CI [2.50, 4.34], p < .001\%\)) for males
only. The CHOCO model revealed that attractiveness had a significantly
positive linear relationship with the probability \emph{p} of judging
faces as real (STATS), but a quadratic relationship with the confidence
in real judgments (\emph{confright}; STATS). Attractiveness was also
associated with less confidence in fake judgments (\emph{confleft};
STATS), a quadratic relationship with left precision (\emph{precleft};
STATS), and related linearly with a stronger bias towards extreme Real
responses (\emph{bex}; STATS). These effects were present for males
only, and no relationship was found for females. The reliability of the
effect of attractiveness (the random slopes) was very low for all
parameters (D-vour \textless{} 0.01). The effect of Beauty was the same
as of attractiveness.

>>>>

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, rightrule=.15mm, bottomrule=.15mm, colback=white, arc=.35mm, breakable, toprule=.15mm, leftrule=.75mm, opacityback=0, left=2mm]

@@

Despite significant advancements in psychological science following the
replication crisis (\citeproc{ref-open2015estimating}{Collaboration,
2015}), its progress is still hindered by its sub-optimal (or
inappropriate) usage of statistical tools
(\citeproc{ref-makowski2023we}{Makowski \& Waggoner, 2023}). A prevalent
issue is the continued reliance on linear models that assume normally
distributed (Gaussian) data\footnote{More specifically, that the outcome
  is distributed according to a Normal distribution which parameters are
  expressed as a linear function of the predictors.} - as this
assumption often does not hold true for many types of psychological
outcomes. For instance, reaction times typically exhibit skewed
distributions, choices can be represented as binary variables, and count
data consists of strictly positive integers. Applying models that
presume normality and model the ``mean'' of the outcome variable can
lead to misinterpretation and potentially misleading conclusions when
applied indiscriminately. It is thus important that psychologists use
models that can best describe (or generate) the data they collect, to
fully exploit them and bring more nuance and accuracy to their
conclusions.

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./figures/fig1.png}

@@

Decision-making is often conceptualized as involving distinct processes:
the choice itself, and the confidence associated with that choice. In
experimental paradigms, these can be somewhat disentangled by prompting
participants to make a discrete choice selection (e.g., ``True''
vs.~``False''), followed by a separate confidence rating. However, this
artificial separation makes its joint analysis difficulty, and may not
reflect real-world decision-making, where individuals often express both
choice and confidence simultaneously using a single, continuous scale.
In such scales, each side can represent a distinct latent category, and
the distance from the midpoint can indicates the level of confidence or
certainty. This integrated response format typically results in bimodal
distributions, with peaks corresponding to the mean confidence on either
side. Traditional beta regression models, which assume unimodal
distributions within the (0,1) interval, are ill-suited for such data.
One alternative is to transform the data into two variables a
posteriori: binarizing the side to represent choice and calculating the
absolute distance from the midpoint to represent confidence. These can
then be modeled separately, for instance, using logistic regression for
choice and beta regression for confidence (see
\citeproc{ref-makowski2025too}{Makowski, Te, et al., 2025} for an
example). While this approach can provide additional insights into
underlying mechanisms compared to a unique model, it assumes
psychological and statistical independence between choice and
confidence, which may not hold true in practice.

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./figures/fig2.png}

@@

\textbf{Study 1} aims to evaluate the CHOCO model's ability to better
capture subjective scale responses that (potentially) reflect an
underlying discrete choice, in comparison to existing models such as the
ZOIB and Beta-Gate. Specifically, we will assess whether 1) CHOCO
provides improved model fit, 2) yields deeper insights into
population-level effects than traditional approaches (gender differences
in reality beliefs), and 3) allows for the reliable estimation of
interpretable individual-level parameters through random effects.
\textbf{Study 2} will apply this model to more subtle effects, such as
the effect perceived facial attractiveness on reality judgments, and
test the ability to fit alternative data structures (scales with ordinal
response options and mid-points). To this end, we analyze data from two
separate studies in which participants judged whether a face image was
AI-generated (``fake'') or a real photograph.

@@

In today's post-truth era, the proliferation of advanced AI technologies
has made it increasingly challenging to distinguish between authentic
and synthetic media, bearing significant implications for information
integrity and public trust
(\citeproc{ref-lewandowsky2017beyond}{Lewandowsky et al., 2017}). As
traditional cues become less reliable (e.g., visual glitches and
artefacts in generated images; formulaic generated text, etc.), people
increasingly depend on contextual information and cognitive heuristics
to assess authenticity, a process referred to as ``simulation
monitoring'' (\citeproc{ref-makowski2019phenomenal}{Makowski, Sperduti,
et al., 2019}).

This reliance on alternative epistemological sources is particularly
pronounced under conditions of high ambiguity, where the
decontextualization of information, especially prevalent in online
environments, complicates authenticity assessments. An open question in
this domain is the extent to which reality judgments are influenced by
the stimuli themselves versus stable individual characteristics like
personality, expectations or expertise - or transient
psychophysiological states (\citeproc{ref-makowski2025too}{Makowski, Te,
et al., 2025}).

Images of faces - socially and perceptually rich stimuli for which
AI-generation has been particularly successful - are a paradigmatic
example that have been used to investigate reality judgments
(\citeproc{ref-azevedo2020body}{Azevedo et al., 2020};
\citeproc{ref-makowski2025too}{Makowski, Te, et al., 2025};
\citeproc{ref-nightingale2022ai}{Nightingale \& Farid, 2022};
\citeproc{ref-tucciarelli2022realness}{Tucciarelli et al., 2022}).
Studies asking participants to judge whether face images are real or
artificially generated reveal that such judgments can be shaped by
low-level features (e.g., clarity, symmetry), higher-level attributes
(e.g., attractiveness, trustworthiness), and interindividual
variability. In the present study, we apply the CHOCO model to such data
to evaluate its capacity to recover interpretable parameters related to
individual-level determinants of reality beliefs.

@@

Using the open-accss data from Makowski, Te, et al.
(\citeproc{ref-makowski2025too}{2025}), we included all heterosexual and
bisexual (as these two groups did not seem to differ based on
preliminary analyses and were thus grouped to maximize power) male and
female participants, for a final sample of 141 participants (Mean age =
28.4, SD = 9.0, range: {[}19, 66{]}; Sex: 47.5\% females). For each
participant, we included only stimuli of the opposite gender (i.e., all
89 female faces for men and 20 male faces for women).

@@

In the first phase, participants viewed 109 neutral-expression
photographs of faces (random order, display time of 3 s) from the
American Multiracial Face Database (AMFD,
\citeproc{ref-chen2021broadening}{Chen et al., 2021}). After each image,
participants rated the face on trustworthiness, familiarity,
attractiveness, and beauty using visual analog scales. In the second
phase, participants were informed that ``about half of the previously
seen images were AI-generated''. The same faces were presented again in
a new random order (same display time), followed by ratings of
``reality'' (whether they believed the image was fake - left anchor - or
real - right anchor).

@@

We fitted 3 models to predict the reality ratings: a ZOIB model, a
Beta-Gate model, and the CHOCO model. For all models and each parameter,
the full formula was entered:
\(Real\sim Sex + (1 | Participant) + (1 | Item)\) (with Sex as the main
predictor and participants and items entered as random intercepts). The
models were run using \emph{brms}
(\citeproc{ref-burkner2017brms}{Bürkner, 2017}) R package, and analyzed
using the \emph{easystats} collection of packages
(\citeproc{ref-ludecke2020extracting}{Lüdecke et al., 2020};
\citeproc{ref-makowski2025modelbased}{Makowski, Ben-Shachar, et al.,
2025}; \citeproc{ref-patil2022datawizard}{Patil et al., 2022}). To
maximize the comparability across models We used the default priors
(uniform) for all models, and we ran 16 chains of 1400 iterations each
on the University of Sussex High-Performance Computing (HPC) cluster.

@@

Finally, we will run a correlation analysis of the models'
individual-level estimates against ``empirical'' (indices computed
directly on the observed data), including the empirical \emph{p}, the
overall \emph{conf}, \emph{pex} and \emph{bex} (respectively calculated
as \(P(y > 0.5)\); \(mean(|y - 0.5|)\); \(P(y \in [0, 1])\); and
\(P(y == 1) / P(y \in [0, 1])\)), assessing whether the model's estimate
are in-line with easily interpretable indices.

@@

{\caption{{Top: Model comparison revealed that the CHOCO model was a
significantly better fit for the data (raw distribution in purple)
compared to a ZOIB or Beta-Gate models, capturing its bimodal
distribution. Bottom: the CHOCO model can be set to estimate effect on
any of its parameters, such as the overall probability of responding on
one side as well as the confidence in both choices. We illustrate this
by showing the effect of Sex on the distribution of reality beliefs
based on a CHOCO model.}{\label{fig-three}}}}

@@

The models did converge without divergent transitions, and the effective
sample size was sufficient for all parameters (all
\(n_{\text{eff}} > 1000\)). The difference in predictive accuracy, as
indexed by Expected Log Predictive Density (ELPD-WAIC), suggests that
the \emph{CHOCO} is the best model (\(ELPD = -203.54\)), followed by
\emph{Beta-Gate} (\(\Delta_{ELPD} = -1794.57 \pm 63.12, ~p < .001)\) and
\emph{ZOIB} (\(\Delta_{ELPD} = -1833.59 \pm 63.52, ~p < .001\)). See
Figure~\ref{fig-three} for the posterior predictive checks, showing that
only the CHOCO model managed to capture the bimodal distribution of
data.

@@

The ZOIB model suggested that women had higher mean scores of reality
beliefs
(\(\mu_{~\text{Female}} = 0.20,~95\%~ CI~[0.03, 0.37], ~pd = 98.77\%\)),
less extreme values
(\(zoi_{~\text{Female}} = -1.24,~95\%~ CI~[-2.46, -0.07], ~pd = 98.25\%\))
but more ones relative to zeros
(\(coi_{~\text{Female}} = 1.63,~95\%~ CI~[0.70, 2.56], ~pd = 99.97\%\)).

The Beta-Gate model similarly suggested that women had higher mean
scores of reality beliefs
(\(\mu_{~\text{Female}} = 0.21,~95\%~ CI~[0.02, 0.40], ~pd = 98.52\%\)),
less extreme values
(\(pex_{~\text{Female}} = -1.34,~95\%~ CI~[-2.56, -0.17], ~pd = 98.75\%\)),
and a greater tendency to answer one relative to zero
(\(bex_{~\text{Female}} = 1.13,~95\%~ CI~[0.34, 1.91], ~pd = 99.76\%\)).

The CHOCO model shows that women had a higher probability \emph{p} of
judging faces as real
(\(P_{~\text{Female}} = 0.45,~95\%~ CI~[0.05, 0.86], ~pd = 98.54\%\)),
but are not more confident when doing so
(\(confright_{~\text{Female}} = -0.13,~95\%~ CI~[-0.47, 0.21], ~pd = 77.73\%\)).
However, they were less confident when answering that an image was
AI-generated
(\(confleft_{~\text{Female}} = -0.53,~95\%~ CI~[-0.89, -0.17], ~pd = 99.84\%\)).
There were also less likely to produce extreme answers
(\(pex_{~\text{Female}} = -1.19,~95\%~ CI~[-2.40, 0.01], ~pd = 97.64\%\)),
but no strong evidence supporting a directional bias was observed
(\(bex_{~\text{Female}} = 0.48,~95\%~ CI~[-0.12, 1.09], ~pd = 94.32\%\)).

@@

{\caption{{Top: the participant and item-level estimates for the CHOCO
parameters, under the distribution of their point-estimates. Reliable
effects are characterised by a higher between-group variability relative
(the dispersion of the distribution of point-estimates) to its
within-group variability (the average uncertainty of individual
estimates represented by the error bar). For example, inter-individual
variability in the parameters reflecting the confidence in left and
right choices is reliably capture, as opposed to the inter-individual
variability in the `bex' parameter. Bottom: correlation matrix between
participant-level estimates from different models and empirical indices
(e.g., the raw proportion of responses on the right per participant).
The CHOCO parameters are easily interpretable, strongly correlating with
their empirical counterparts.}{\label{fig-four}}}}

\includegraphics[width=0.95\linewidth,height=\textheight,keepaspectratio]{./figures/fig4.png}

@@

The CHOCO model yielded reliable estimates (Figure~\ref{fig-four}) for
all parameters except \emph{bex} for participants (\emph{confright}
D-vour = 0.94, \emph{confleft} D-vour = 0.91, \emph{pex} D-vour = 0.79,
\emph{p} D-vour = 0.73, \emph{precright} D-vour = 0.77, \emph{precleft}
D-vour = 0.67). Item's variability was primarily reflected through the
\emph{p} parameter (D-vour = 0.86).

@@

Finally, we also show that the CHOCO model was able to capture reliable
and interpretable individual-level parameters, supporting its value to
measure inter-individual differences. An interesting dissociation
emerged between participant- and item-level variability: the latter
seemed mostly to be represented in the \emph{p} parameter, while
participants reliably varied in most of the components (aside from
\emph{bex}). This could suggest that external item characteristics
primarily influence the probability of being judged as real vs.~fake,
while the expressed confidence is first and foremost an individual
characteristic.

@@

In this evolving landscape, understanding the cognitive mechanisms that
underpin our judgments of reality becomes paramount. Despite the
increasing prevalence of digitally altered or AI-generated content,
humans still rely on certain heuristics to assess authenticity. One such
heuristic might be facial attractiveness. Attractiveness appraisals are
known to be automatic and unconscious (\citeproc{ref-hou2023neural}{Hou
et al., 2023}; \citeproc{ref-hung2016unconscious}{Hung et al., 2016};
\citeproc{ref-luo2019robust}{Luo et al., 2019}), and carry strong
real-life consequences, as demonstrated by the large body of literature
on the ``beauty premium''\footnote{As an eloquent example, Monk Jr et
  al. (\citeproc{ref-monk2021beholding}{2021}) reported in a large
  representative US sample that the magnitude of earnings disparities
  among white women along the perceived attractiveness continuum exceeds
  in magnitude the canonical black-white race gap.}
(\citeproc{ref-gulati2024beautiful}{Gulati et al., 2024};
\citeproc{ref-kukkonen2024beauty}{Kukkonen et al., 2024};
\citeproc{ref-little2021facial}{Little, 2021};
\citeproc{ref-pandey2021face}{Pandey \& Zayas, 2021}).

Although its role as a potential modulator of reality beliefs remains
under-explored, Makowski, Te, et al.
(\citeproc{ref-makowski2025too}{2025}) found significant associations
between participants' realness ratings and facial attractiveness, with a
potential sexual dymorphism: male participants judged attractive faces
as being likely more real, whereas evidence suggested a milder and
quadratic (U-shaped) relationship between attractiveness and reality
beliefs for females. These findings offer a complementary perspective to
those of Miller et al. (\citeproc{ref-miller2023ai}{2023}), who reported
that participants used attractiveness as a distinguishing cue between
real and AI-generated faces. These results highlight a possible
bidirectional and context-dependent influence of attractiveness on
reality judgments, of which the exact shape needs further investigation.

@@

The first sample includes the same participants as study 1 (N = 141, see
above). For the second sample (data and preprocessing information
available at https://github.com/RealityBending/FakeFace2), 248
participants were initially recruited through academic platforms (SONA
and SurveySwap). We removed 15 participants with data suggesting
low-effort responding as well as those (N=16) that did not believe in
the experimental manipulation and were fully confident that all images
were real or fake. Similarly to the first sample, we included all hetero
and bisexual participants and stimuli from the opposite gender,
resulting in a final sample included 189 participants (Mean age = 28.4,
SD = 14.0, range: {[}18, 69{]}; Sex: 76.2\% females). The study was
approved by the University of Sussex Ethics Board (ER/ST633/1).

@@

For the second sample, the procedure was relatively similar to that of
the first sample (described above in Study 1) with a few key
differences.

The main difference was the introduction of an experimental
manipulation: while for sample 1, participants were simply informed of
the presence of AI-generated stimuli among photographs (not providing
information as to specifically which image), the reality beliefs were
directly manipulated in sample 2. At the beginning of the experiment, a
cover story presented the study as a partnership with an AI startup
aimed at testing the quality of a new face AI-generation algorithm.
Following that, participants would see the 109 neutral-expression
photographs from the AMFD database (1 s), each preceded by a randomly
assigned textual cue indicating whether the image was ``AI-generated''
or ``Photograph'' (2 s). Ratings of attractiveness, beauty and
trustworthiness were collected after each image.

This phase of the experiment concluded with multi-choice questions
asking participants to indicate whether they believed in the cover
story. The second phase would start with a new set of instructions
(falsely) revealing that the cues were ``mixed up'' (shuffled randomly),
and that they would now be presented with the faces again (1 s) followed
by an assessment of their own beliefs about whether the image was real
or fake.

The second main difference was the subjective ratings' format, collected
using a 7-point Likert scale ranging from 0 to 6 (which included a clear
midpoint option), rather than a visual analog scale for sample 1. The 3
buttons on the left side (0, 1, 2) were colored in red and corresponded
to the more-or-less pronounced belief that the image was
``AI-generated'', the 3 values on the right (4, 5, 6) for ``Photograph''
were colored in green, and the middle value (3), representing an
undecided option, was colored in yellow.

@@

To compare the benefits of CHOCO models to a ``traditional'' analytic
approach, we started by fitting a frequentist linear mixed model to
predict reality beliefs with the formula
\(Real\sim Sex / poly(Attractive, 2) + (poly(Attractive, 2)|Participant) + (1|Item)\).
The second degree orthogonal polynomial term was included to allow for
potentially non-linear relationships (note that the first and second
degree effects of orthogonal polynomials can be interpreted
independently as the linear part and the ``curvy'' part of the
relationship). For the CHOCO model, mildly informative and
effect-agnostic (i.e., centered at zero) priors were used. The same
formula was used for all parameters, except that items were only
included as random effects for \emph{p}, and participants were not
included for \emph{bex} (based on the reliability analysis of Study 1).

For sample 2, the analysis was based on that of Sample 1. The main
differences are 1) the inclusion of the ``Condition'' (whether the
picture was presented as ``Real'' of ``Fake'' in the first phase of the
experiment) as an additional predictor (entered as the only random slope
for all participant random effects); 2) the inclusion of an additional
parameters, \emph{pmid}, modelling the probability of answering the
middle-point of the Likert scale (representing an ``undecided'' option);
3) as they were effectively only 3 distinct values on each side of the
scale, and to showcase the flexibility of the CHOCO model, we decided to
\emph{not} treat the extreme values as 0 and 1 (and model them via the
separate parameters \emph{pex} and \emph{bex}, which would distributions
to be estimated from only 2 values), but instead to treat them as part
of the continuous distribution. The 7 response options were rescaled to
be evenly spaced between 0 and 1 excluded (i.e., {[}0.125, 0.875{]}).
The \emph{pex} and \emph{bex} parameters were fixed to 0, making for a
slightly more parsimonious model.

@@

{\caption{{The effect of attractiveness on reality beliefs. Top: the
effect of different levels of attractiveness (shades of red) on the
CHOCO distribution of the reality ratings in both samples, showing
primarily that, for male participants, more attractive faces were judged
more likely as a photograph rather than an AI-generated image. For the
second sample, stimuli that were during the initial viewing presented as
`AI-generated' are represented via a dotted line. Bottom: The impact of
attractiveness (x-axis) on different CHOCO distribution parameters, for
male and female participants (the AI-generated condition is added as a
transparent line for sample 2). Given the presence of polynomial terms,
significant effects are denoted by a symbol representing the shape of
the relationship (/ or \textbackslash{} for linear, U or inverse-U for
quadratic links).}{\label{fig-five}}}}

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./figures/fig5.png}
